Variance reduction for MCMC methods via martingale representations

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
update vr

\newcommand{\eric}[1]{\todo[color=purple!10, author=Eric]{#1}}
\newcommand{\erici}[1]{\todo[inline,color=purple!10, author=Eric]{#1}}
\newcommand{\denis}[1]{\todo[color=green!10, author=Denis]{#1}}
\newcommand{\denisi}[1]{\todo[inline,color=green!10, author=Denis]{#1}}
\newcommand{\sergey}[1]{\todo[color=blue!10, author=Sergey]{#1}}
\newcommand{\sergeyi}[1]{\todo[inline,color=red!10, author=Sergey]{#1}}

@@

\section{Setup}\label{sec:setup}
Let  \(\mathcal{X}\) be a domain in \( \mathbb{R}^d.\)  Our aim is to numerically compute  expectations of the form
\[
\pi(f)=\int_{\mathcal{X}} f(x)\pi(\rmd x),
\]
where \(f:\) \(\mathcal{X}\longrightarrow \mathbb{R}\) and \(\pi\) is a probability measure supported on \(\mathcal{X}.\)
If  the dimension of the space \(\mathcal{X}\) is large and \(\pi(f)\) can not be computed analytically, one can apply Monte Carlo methods. However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case. One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC), where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p\geq 0}\) such that \(\pi\) is its unique invariant measure. In this paper we study a class of MCMC algorithms with \((X_p)_{p\geq 0}\) satisfying the  the following recurrence relation:
\begin{equation}
\label{eq:chain_gen}
X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x_0,
\end{equation}
for some i.i.d.  random vectors \(\xi_p\in \mathbb{R}^m\) with distribution \(P_{\xi}\)
and some Borel-measurable
functions $\Phi_{p}\colon\mathcal{X}\times\mathbb{R}^{m}\to\mathcal{X}.$
In fact, this is quite general class of Markov chains (see Theorem~1.3.6 in ~\cite{moulines2018})
and many well-known MCMC algorithms can be represented in form \eqref{eq:chain_gen}.
Let us consider two popular examples.
\begin{example}[Unadjusted Langevin Algorithm]
\label{exam:langevin-algorithm}
Fix a sequence of positive time steps \((\gamma_p)_{p\geq 1}.\) Given a Borel function $\mu\colon\mathbb{R}^{d}\to\mathbb{R}^{d}$,
consider a non-homogeneous
discrete-time Markov chain $(X_{p})_{p\geq0}$ defined by
\begin{equation}\label{eq:chain}
X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
standard Gaussian random vectors. If
$\mu= \nabla U$
for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
\begin{equation}\label{eq:stationary_distr}
\pi(x)= \mathrm{const}\,\rme^{-\frac{U(x)}{2}}, \quad \mathrm{const}= \left. 1 \middle /{\int_{\mathbb{R}^{d}} \rme^{-\frac{U(x)}{2}}\, \rmd x,} \right.
\end{equation}
provided that \(\int_{\mathbb{R}^{d}} \rme^{-U(x)}\,\rmd x\) is finite. This method is usually referred to as
Unadjusted Langevin Algorithm (ULA).
In fact the Markov chain~\eqref{eq:chain}
arises as the Euler-Maruyama discretization
of the Langevin diffusion
\[
\rmd Y_t=-\mu(Y_t)\,\rmd t+ \rmd W_t
\]
with nonnegative time steps $(\gamma_p)_{p\ge1}$,
and,  under mild technical conditions, the latter Langevin diffusion admits $\pi$
of~\eqref{eq:stationary_distr}
as its unique invariant distribution; see \cite{dalalyan2017theoretical} and \cite{durmus:moulines:2017}.
\end{example}
\begin{example}[Metropolis-Adjusted Langevin Algorithm]
The Metropolis-Hastings algorithm
associated with a target density \(\pi\) requires the choice of a sequence of conditional densities  \((q_p)_{p\geq 1}\) also called proposal or candidate kernels. The transition from the value  of the Markov chain \(X_p\)  at time \(p\)
and its value at time \(p + 1\) proceeds via the following transition step:

@@

$\phi_k(\xi)$, $k\ge1$, are centered. As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$. We denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration generated by independent random variables $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$. For $l \le p$ and $x \in \rset^d$, put for $x,y \in \rset^d$ functions $\Phi_k(x,y)=x-\gamma_k\mu(x)+y$. Then we can define
X^x_{l,p}=G_{l,p}(x,\sqrt{\gamma_l}\xi_{l},\ldots,\sqrt{\gamma_p}\xi_{p}),
with the function \(G_{l,p}:\) \(\mathbb{R}^{d\times(p-l+2)}\to \mathbb{R}^{d}\) is defined as
G_{l,p}(x,y_l,\ldots,y_p):=\Phi_p(\cdot,y_{p})\circ\Phi_{p-1}(\cdot,y_{p-1})\circ\ldots\circ\Phi_{l}(x,y_{l})
For simplicity we will write $X^x_{p}$ and $G_{p}$ instead of $X^x_{0,p}$ and $G_{0,p}$. As a consequence, for a function $f\colon \rset^d \to \rset$ we have
f\left(X^x_{p}\right) =%f\circ G_{p}(x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}).
f\circ G_{p}(x,\sqrt{\gamma}\xi_{1},\ldots,\sqrt{\gamma}\xi_{p})
%\erici{It would be perhaps interesting to state the conditions satisfied by the functions $(\Phi_l)_{l \geq 1}$ here}

@@

 Let $f$ be a Borel function $\rset^{d}\to\mathbb R$ such that
$\mathsf{E}\left[\left |f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some sequence of functions $(\Phi_l)_{l\geq 1}.$

@@

%a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
a_{p,l,k}(x)=\mathsf{E}\left[f(X^x_{l-1,p})\phi_k\left(\xi_{l}\right)\right], \quad p\geq l, \quad k\in \mathbb{N}

@@

\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
The expansion obviously holds for $p=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), we have for all $x \in \rset^d$,
\eric{it is clear, but we must say in which sense this equality should be understood}
\[
f(X_{1})=\PE[f(X_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
\]
with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$,
provided $\PE[|f(X_{1})|^{2}]<\infty.$
\tcr{Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$
$l=1,2,\ldots, $
and $\mathcal{G}_0=\mathrm{triv}$.}
\eric{do we need this, this is recalled ten lines above}
Suppose that (\ref{eq:mart_repr})
holds for $p=q$, all $j<q$, and all Borel-measurable functions
$f$ with $\mathsf{E}\left[|f(X_{q})|^2\right]<\infty$.
Let us prove it for $p=q+1$.
Given $f$ with $\mathsf{E}\left[|f(X_{p})|^2\right]<\infty$,
due to the orthonormality and completeness
of the system $\left(\phi_{k}\right)$, we get by
conditioning on $\mathcal{G}_{q}$,
\[
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{p,q+1,k}\phi_{k}(\xi_{q+1}),
\]
where
\begin{equation*}
\alpha_{p,q+1,k}
=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
\end{equation*}
By the Markov property of $(X_{l})$,
we have
$\mathsf{E}[f(X_{p})|\mathcal{G}_q]
=\mathsf{E}[f(X_{p})|X_{q}]$.
Furthermore, a calculation involving
intermediate conditioning on $\mathcal G_{q+1}$
and the recurrence relation
$X_{q+1}=\Phi_{q+1}(X_{q},\xi_{q+1})$
verifies that
\[
\alpha_{p,q+1,k}=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|X_{q}\right]
=a_{p,q+1,k}(X_{q})
\]
for suitably chosen
Borel-measurable functions $a_{p,q+1,k}$.
We thus arrive at
\begin{align}
\label{berm:sig_X}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]+\sum_{k\geq1}a_{p,q+1,k}(X_{q})\phi_{k}(\xi_{q+1}),
\end{align}
which is the required statement in the case $j=q$.
Now assume $j<q$.
The random variable
$\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]$
is square integrable and has the form
$g(X_{q})$,
hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]=\mathsf{E}\left[\left.f(X_{p})\right|X_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{p,l,k}(X_{l-1})\phi_{k}(\xi_{l})
\end{equation}
with
\begin{align*}
a_{p,l,k}(X_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X_{p})\right|\mathcal{G}_{q}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
= \mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]\\
&=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|X_{l-1}\right].
\end{align*}
Formulas \eqref{berm:sig_X}
and~\eqref{eq:28082017a1} conclude the proof.
\end{proof}

\begin{cor}
If  all the kernels $\Phi_l,$ $l\geq 1,$ are equal,  then  the representation \eqref{eq:mart_repr} takes the form
\begin{eqnarray*}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X_{l-1})\phi_k\left(\xi_{l}\right)
\end{eqnarray*}
with
\begin{eqnarray*}
\bar a_{r,k}(x)=\mathsf{E}\left[f(X^x_{r})\phi_k\left(\xi_{1}\right)\right], \quad r, k\in \mathbb{N}.
\end{eqnarray*}
\end{cor}

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
The coefficients \(a_{p,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{p,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{p,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{p,l}(x)=\mathsf{E}\left[f(X^x_{l,p})\right],\) \(p\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\ldots,p-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{p,l}(x)=\mathsf{E}\left[Q_{p,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
In the case $\Phi_l=\Phi$ for all $l\geq 1,$ we have
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
\end{equation}
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the case of time homogeneous transition kernels and introduce a weighted average estimator $\pi_{n}^{N}(f)$ of the form
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X_{p}),
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{eqnarray}
\nonumber
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
\\
&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right),
\label{eq:29032018a5}
\end{eqnarray}
where
\begin{equation}
\label{eq:definition-A-s,k}
A_{s,k}(x)=\frac{1}{s}\sum_{r=1}^{s}\bar a_{r,k}(x).
\end{equation}
Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\)}, we obtain 
\[\mathsf{E}[A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X_{l-1})\mathsf{E}[\left.\phi_{k}\left(\xi_{l}\right)\right|\mathcal{G}_{l-1}]] = 0\]
\eric{I do not understand this} and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\begin{equation}
\label{eq:expression-variance}
\PVar[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X_{N+l-1})]
\end{equation}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.

\sergey{N and N0 were defined here before}
\eric{these notations are used before; it is best to put all the notations in a short paragraph just after the introduction}

>>>>

\newcommand{\eric}[1]{\todo[color=purple!10, author=Eric]{#1}}
\newcommand{\erici}[1]{\todo[inline,color=purple!10, author=Eric]{#1}}
\newcommand{\denis}[1]{\todo[color=green!10, author=Denis]{#1}}
\newcommand{\denisi}[1]{\todo[inline,color=green!10, author=Denis]{#1}}
\newcommand{\sergey}[1]{\todo[color=red!10, author=Sergey]{#1}}
\newcommand{\sergeyi}[1]{\todo[inline,color=red!10, author=Sergey]{#1}}

@@

\section{Setup}\label{sec:setup}
Let  \(\mathcal{X}\) be a domain in \( \mathbb{R}^d.\)  Our aim is to numerically compute  expectations of the form
\[
\pi(f)=\int_{\mathcal{X}} f(x)\pi(\rmd x),
\]
where \(f:\) \(\mathcal{X}\longrightarrow \mathbb{R}\) and \(\pi\) is a probability measure supported on \(\mathcal{X}.\)
If  the dimension of the space \(\mathcal{X}\) is large and \(\pi(f)\) can not be computed analytically, one can apply Monte Carlo methods. However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case. One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC), where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p\geq 0}\) such that \(\pi\) is its unique invariant measure. In this paper we study a class of MCMC algorithms with \((X_p)_{p\geq 0}\) satisfying the  the following recurrence relation:
\begin{equation}
\label{eq:chain_gen}
X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x_0,
\end{equation}
for some i.i.d.  random vectors \(\xi_p\in \mathbb{R}^m\) with distribution \(P_{\xi}\)
and some Borel-measurable
functions $\Phi_{p}\colon\mathcal{X}\times\mathbb{R}^{m}\to\mathcal{X}.$
In fact, this is quite general class of Markov chains (see Theorem~1.3.6 in ~\cite{moulines2018})
and many well-known MCMC algorithms can be represented in form \eqref{eq:chain_gen}.
Let us consider two popular examples.
\begin{example}[Unadjusted Langevin Algorithm]
\label{exam:langevin-algorithm}
Fix a sequence of positive time steps \((\gamma_p)_{p\geq 1}.\) Given a Borel function $\mu\colon\mathbb{R}^{d}\to\mathbb{R}^{d}$,
consider a non-homogeneous
discrete-time Markov chain $(X_{p})_{p\geq0}$ defined by
\begin{equation}\label{eq:chain}
X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{2 \gamma_{p+1}}Z_{p+1},\end{equation}
where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
standard Gaussian random vectors. If
$\mu= \nabla U$
for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
\begin{equation}\label{eq:stationary_distr}
\pi(x)= \mathrm{const}\,\rme^{-U(x)}, \quad \mathrm{const}= \left. 1 \middle /{\int_{\mathbb{R}^{d}} \rme^{-U(x)}\, \rmd x,} \right.
\end{equation}
provided that \(\int_{\mathbb{R}^{d}} \rme^{-U(x)}\,\rmd x\) is finite. This method is usually referred to as
Unadjusted Langevin Algorithm (ULA).
In fact the Markov chain~\eqref{eq:chain}
arises as the Euler-Maruyama discretization
of the Langevin diffusion
\[
\rmd Y_t=-\mu(Y_t)\,\rmd t+ \rmd W_t
\]
with nonnegative time steps $(\gamma_p)_{p\ge1}$,
and,  under mild technical conditions, the latter Langevin diffusion admits $\pi$
of~\eqref{eq:stationary_distr}
as its unique invariant distribution; see \cite{dalalyan2017theoretical} and \cite{durmus:moulines:2017}.
\end{example}
\begin{example}[Metropolis-Adjusted Langevin Algorithm]
The Metropolis-Hastings algorithm
associated with a target density \(\pi\) requires the choice of a sequence of conditional densities  \((q_p)_{p\geq 1}\) also called proposal or candidate kernels. The transition from the value  of the Markov chain \(X_p\)  at time \(p\)
and its value at time \(p + 1\) proceeds via the following transition step:

@@

$\phi_k(\xi)$, $k\ge1$, are centered. As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) In the sequel, we denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration
generated by generated by $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
\erici{It would be perhaps interesting to state the conditions satisfied by the functions $(\Phi_l)_{l \geq 1}$ here}

@@

 Let $f$ be a Borel function $\mathbb{R}^{d}\to\mathbb R$ such that
$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some sequence of functions $(\Phi_l)_{l\geq 1}.$

@@

a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.

@@

\begin{cor}
If  all the kernels $\Phi_l,$ $l\geq 1,$ are equal,  then  the representation \eqref{eq:mart_repr} takes the form
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X_{l-1})\phi_k\left(\xi_{l}\right)
with
\bar a_{r,k}(x)=\mathsf{E}\left[\left.f(X_{r})\phi_k\left(\xi_{1}\right)\right|X_{0}=x\right], \quad r, k\in \mathbb{N}.

\end{cor}
\begin{proof}[Proof of Theorem~\ref{prop:29032018a1}]
The expansion obviously holds for $p=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system of , we have for all $x \in \rset^d$,
\eric{it is clear, but we must say in which sense this equality should be understood}
f(X^x_{1})=\PE[f(X_{1}^x)]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$,
provided $\PE[|f(X^x_{1})|^{2}]<\infty.$
\tcr{Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$
$l=1,2,\ldots, $
and $\mathcal{G}_0=\mathrm{triv}$.}
\eric{do we need this, this is recalled ten lines above}
Suppose that (\ref{eq:mart_repr})
holds for $p=q$, all $j<q$, and all Borel-measurable functions
$f$ with $\mathsf{E}\left[|f(X_{q})|^2\right]<\infty$.
Let us prove it for $p=q+1$.
Given $f$ with $\mathsf{E}\left[|f(X_{p})|^2\right]<\infty$,
due to the orthonormality and completeness
of the system $\left(\phi_{k}\right)$, we get by
conditioning on $\mathcal{G}_{q}$,
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{p,q+1,k}\phi_{k}(\xi_{q+1}),
\alpha_{p,q+1,k}
=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
By the Markov property of $(X_{l})$,
we have
$\mathsf{E}[f(X_{p})|\mathcal{G}_q]
=\mathsf{E}[f(X_{p})|X_{q}]$.
Furthermore, a calculation involving
intermediate conditioning on $\mathcal G_{q+1}$
and the recurrence relation
$X_{q+1}=\Phi_{q+1}(X_{q},\xi_{q+1})$
verifies that
\alpha_{p,q+1,k}=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|X_{q}\right]
=a_{p,q+1,k}(X_{q})
for suitably chosen
Borel-measurable functions $a_{p,q+1,k}$.
We thus arrive at
\label{berm:sig_X}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]+\sum_{k\geq1}a_{p,q+1,k}(X_{q})\phi_{k}(\xi_{q+1}),
which is the required statement in the case $j=q$.
Now assume $j<q$.
The random variable
$\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]$
is square integrable and has the form
$g(X_{q})$,
hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]=\mathsf{E}\left[\left.f(X_{p})\right|X_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{p,l,k}(X_{l-1})\phi_{k}(\xi_{l})
with
a_{p,l,k}(X_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X_{p})\right|\mathcal{G}_{q}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
= \mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]\\
&=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|X_{l-1}\right].
Formulas \eqref{berm:sig_X}
and~\eqref{eq:28082017a1} conclude the proof.
From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
The coefficients \(a_{p,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
a_{p,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{p,l}\left(\Phi_l(x,\xi)\right)\right]
with \(Q_{p,l}(x)=\mathsf{E}\left[\left.f(X_{p})\right|X_{l}=x\right],\) \(p\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\ldots,p-1\}$
\label{eq:qpl}
Q_{p,l}(x)=\mathsf{E}\left[\left.Q_{p,l+1}(X_{l+1})\right|X_{l}=x\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
In the case $\Phi_l=\Phi$ for all $l\geq 1,$ we have
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
with  \(Q_{r}(x)=\mathsf{E}\left[\left.f(X_{r})\right|X_{0}=x\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the case of time homogeneous transition kernels and introduce a weighted average estimator $\pi_{n}^{N}(f)$ of the form
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X_{p}),
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right),
\label{eq:29032018a5}
\label{eq:definition-A-s,k}
A_{s,k}(x)=\frac{1}{s}\sum_{r=1}^{s}\bar a_{r,k}(x).
Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0,\)   the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\label{eq:expression-variance}
\PVar[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X_{N+l-1})]
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma=\gamma_1=\gamma_2=\ldots.$
\tcr{We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$.} \eric{these notations are used before; it is best to put all the notations in a short paragraph just after the introduction}
By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}\Bigl(1-\frac{l}{n}\Bigr)A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
with \(\|\mathbf{k}\|=\max_{i} k_i.\)
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L_U < \infty$ such that
|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \rset^d.
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)$  it holds
\ps{D^2U(x)}{x} \geq (m_U/2)\|x\|^2.
\end{description}
First we analyse the biases of the estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f).$
\paragraph{Squared bias:} Due to the martingale transform structure
of $M_{K,n}^N(f)$,
we have
\mathsf E\left[M_{K,n}^N(f)\right]=0,
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.
Notice that this remains true also when we substitute
the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$
